MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-04-25 rebuttal.txt

First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks. We have addressed them in our paper revision. The individual answers are in the following. Minor remarks that were simply fixed in the paper and required no additional comments are omitted.

Reviewer #1:

Q: How is the parallelization for GPUs actually done? In the section MaBoSS.GPU, it is described, why there was no parallelization over the Boolean functions, but not about the rest. In the MPI section, it states, that each thread is assigned to compute a single trajectory at once. I would assume that this is also true for the GPU implementation, this should, however, be made clear.
A: You are right, we included this piece of information into the last paragraph of Simulation section.

Q: The legend in the figures needs to be more specific to be able to interpret the figures better: e.g. how many synthetic models? (are the 19 points the respective models?)
A: yes, the x-axis shows multiple synthetic models of various sizes. We added this to Figure 2.

Q: In Figure1, the High-End NVIDIA TESLA is either slower or as fast as the consumer grade laptop GPU. Is this purely due to the compilation time (as Figure 2 would suggest, that it makes a difference)? 
A: You are right. You can see for yourself by comparing gpu_out_real.csv and gpu_out_real_a100.csv files in the artifact. We are unsure why the compilation takes longer for high-end GPU than for laptop GPU, after all, it is a proprietary piece of software. Perhaps the loading of shared libraries takes a different amount of time for the operating systems... Due to the uncertainty and the fact that the difference is in hundreds of milliseconds, we do not see this as a detail worth explicitly mentioning in the paper.

Q: There is no demonstration of how a user would actually deal with the code - especially since the GPU version is based on a hard-coded implementation of the rules. A demonstrative example would help to get an idea, how a user would actually get started, and how deep he needs to dig into the code - especially for a biological audience, accessability of such tools is an important point
A: Respectfully, it is not true that the GPU version is based on any hard-coded implementation. The compilation of the rules happens during the runtime without any user intervention. We included this note in the second paragraph of the Simulation section.

Reviewer #2:

Q: Figure 3: The x-axis states “Nodes count (log−scale)”, however, it appears to have a linear scale. The linear relationships claimed in the statement “We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).” is not clear from the provided figures, and should be clarified or, for example provide Figure 3 with a linear-linear or log-log scale.
A: Thanks for pointing out the wrong scale in the plot, it is a linear scale. Regarding the linear dependency, it can not be seen from Fig 3 since it shows compilation time percentages rather than the absolute values. Therefore, plotting a linear-linear scale would not show any linear dependency. Nevertheless, the stated linear dependency is indeed there and we observed it from the plots which we did not include in the paper. The general idea why it holds is that increasing the number of nodes by a value of x adds O(x) new lines to a source file which needs to be compiled. We included this explanation in the paper.

### 2024-04-25 Article.tex
some gpu related fixes

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. Using this technique, the Boolean formulas are compiled as functions into a native binary code, which is directly executed by the GPU. As the main advantage, the formulae are encoded in the instructions, preventing unnecessary fetches of the encoded formulas from other memory. At the same time, the compiler may apply a vast spectrum of optimizations on the Boolean formulas, including case analysis and shortcutting, again resulting in faster evaluation.

@@

Also, due to the involved implementation complexity, we avoided optimization of the computation of individual trajectories by splitting the Boolean function evaluation into multiple threads (thus missing the factor of $n$ threads from the asymptotic analysis). While such optimization might alleviate some cache pressure and thus provide significant performance improvements, we leave its exploration to future work.

@@

This approach allows us to avoid storing the states as the keys, and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions. Further, we use several well-known GPU histogram update optimizations to improve the performance, including shared memory privatization and atomic operations.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models. Each model is simulated with 1 million trajectories. The two panels differ by the inclusion of the runtime GPU code compilation, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

@@

Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes). Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.

>>>>

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. In such an approach, the application reads the model files, writes the formulas as functions in CUDA C++, compiles them using the NVIDIA runtime compiler, and finally runs the simulation --- all without a user intervention.

Using this technique, the Boolean formulas are compiled as functions into a native binary code, which is directly executed by the GPU. As the main advantage, the formulae are encoded in the instructions, preventing unnecessary fetches of the encoded formulas from other memory. At the same time, the compiler may apply a vast spectrum of optimizations on the Boolean formulas, including case analysis and shortcutting, again resulting in faster evaluation.

@@

The work distribution was chosen to be one trajectory simulation per GPU thread. Due to the involved implementation complexity, we avoided optimization of the computation of individual trajectories by splitting the Boolean function evaluation into multiple threads (thus missing the factor of $n$ threads from the asymptotic analysis). While such optimization might alleviate some cache pressure and thus provide significant performance improvements, we leave its exploration to future work.

@@

This approach allows us to avoid storing the states as the keys and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions. Further, we use several well-known GPU histogram update optimizations to improve the performance, including shared memory privatization and atomic operations.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models with sizes ranging from $10$ to $1000$ nodes (x-axis). Each model is simulated with 1 million trajectories. The two panels differ by the inclusion of the runtime GPU code compilation, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

@@

Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes), which can be simply explained by the fact that these model properties extend source files which need to be compiled by a linear factor. Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.
